Should I Airbnb My Property?: \n Full Report

### 2017-02-13 abb_Writeup.Rmd
re-org of names

---
title: "Should I Airbnb My Property?"
author: "Andy Krause and Gideon Aschwanden"
date: "January 9, 2017"
output:
  pdf_document: default
  html_document: default
geometry: margin=3.4cm
bibliography: abb.bib
---

@@

To address these questions and provide some guidance to potential and current investors and owners, we use long-term rental data and nightly Airbnb listing information from the Melbourne (Australia) metropolitan area.  Melbourne has a thriving Airbnb market with over 20,000 total properties listed on the portal. We adopt a scenario-based approach in which we compare potential annual revenues for properties under both a short-term rental (STR) and a long-term rental (LTR) situation. Additionally, we break the results down by property types (apartment vs house), product type (ex. 1 bed, 1 bath) and by location (submarket).  Our results shows that the feasibility of a greater return through a short term rental is dependent on property type, product and location.  In answer to whether or not a owner or investor should opt for a short-term leasing strategy: it depends.

@@

Existing research into short-term rental markets is sparse. However, with the arrival of Airbnb, and the controversies surrounding it, the topic is creating interest both in real estate and economics as well as in a wide variety of fields outside of real estate as evidenced by a growth in recent publications and on-going working papers.

@@

We will not, a priori, specify a location or property type as we hypothesize that the preference of a short-term rental over a long-term one will vary by the location and property characteristics.  Additionally, Australia implements a particular tax policy, often termed `negative gearing', that does allow for favourable tax treatment to real estate investors who experience greater costs than revenues on investment properties^[See @fane2004negative for a complete explanation of negative gearing.  However, as the negative gearing policy requires knowledge of the investor's personal income situation, we will not be considering any effects of this policy in our analysis here.

@@

We will begin our analysis by making a comparison of the short and long-term markets at a metro-level.  What we mean by this is that we will treat the entire region as the functional market and will impute rates and rents with a model using all data.  It is likely that different submarkets operate differently, we will explore this hypothesis after first analyzing the full metropolitan region.

@@

@@

First, we bifurcate our data into houses and apartments.  We then rerun the entire analysis described above for the house and the apartment markets separately. Table 5 and Figures 18 (houses) and 19 (apartments) illustrate the metrics for these two sub-markets.  The raw probability data shows that over 33% of apartments are more profitable in short term leasing as opposed to just 22% of houses.  The two figures show the quantile preference plots both in the raw and with the SVM categorization.  As is evident from these plots, the likely of short-term rental being more profitable is much greater in the apartment market than in the house market.  For apartments, a 60th rate and a 60th percentile occupancy rate has a better than 50% chance of being more profitable in a short term leasing strategy than a long term one.  In fact, so long as an apartment is in the top 1/4 of occupancy rates, it is more profitable in a short term strategy at almost any price point.  Houses, on the other hand, need either very high rates or very high occupancies to justify a short-term rental.

@@

We break the data down further by geographic submarket and property type and compare the results (quantile) in Table 6.  In the Actual (observed) revenues, the probabilty of short-term preference is highest in the core area of the city and lowest in the city, beach and suburban areas.  Interestingly, the rural area has a higher probably than all other but the core.  This is likely explained by the fact that there are few rural properties and those that are show some clustering near rural tourists activities like the Yarra Valley wine region (Figure 8).

In the core and the city, apartments offer a higher probability than homes, the reverse in true in the suburban and rural areas. In the lower density areas it appears that tourists may prefer larger and more spacious dwellings and are willing to pay a premium for them, compared to long-term renters who may experience long commutes from these fringe locations.

Looking at the 'fitted' and 'svm area' figures we see a trend that these svm figures are higher than the actual observed for the core apartment and lower for nearly all of the rest of the categories.  This findings shows that for all but the core apartments, those properties that are more profitable as short-term cluster tightly in the area (the combination of nightly rate and occupancy rate) that are preferable for a short-term strategy. In other words, there is well defined combination or situation in which these properties can be more profitable than in a long-term lease.  The reverse is true of the city core apartments. Here, the area of the plots occupied by short term preference is larger than the total proporation, suggesting that a wide variety of combinations of nightly rates and occupancy rates will result in a short term preference.

@@

Finally, in Model 3 we add in host behavior or policies. Allowing a greater number of guests per bedroom results in a higher probability of short-term preference.  This variable may be operating a proxy for additional sleeping spaces which allow for a higher nightly rate, ceteris paribus, to be charged.  The greater the minimum stay, the lower the probability of short-term preference.  Longer minimum stay may equate to lower occupancy rates as one or two night travelers are exempted from the property.  Owners with longer minimum stays may be looking to minimize costs such as cleaning and administration.  Flexible cancellation policies result in a lower probability of short-term preference, while a Strict policy has a weak, but positive relationship with short-term preference.  Strict policies do help maintain revenues, but may also scare off some potential guests.  Strict cancellation may be a luxury for the best and highly desired properties on the market and, within out model, may be acting as a proxy for some unobserved measure of quality or demand.

@@

Although it gathers considerable interest in both economic and public policy circles, analyses of Airbnb market and their comparison to more traditional long-term leasing strategies remains relatively unaddressed.  This work attempts to fill this void.

The Airbnb market is Melbourne is rapidly expanding, as total listings on the portal have jumped from around 7,000 in 2014 to more than 20,000 in August 2016.  However, of this total, only 6,600 or so of the properties are entire homes or apartments that were listed as available for at least one day during the September 1, 2015 to August 31, 2016 time frame.  Of these, only 4,400 are common bed/bath configurations with non-outlying rates and were booked for at least one night over the period of study.  We caution future researchers and policy analysts to look deeply at initial estimates of Airbnb counts as many of the listings on the market are either room shares and/or not active.

@@

In looking only at the short-term observation with estimated long-term rents we found that approximately one in three Airbnb properties brought in more revenue with its short term strategy than it would have with a long-term approach. As expected, this probability is not equal across space, property type or property size (bed and bath count).  Apartments are much more likely than houses to be more profitable as a short term rental.  Core city properties are more likely than more outlying properties to prefer a short-term strategy, as a larger properties versus smaller ones.  For instance, 45\% of apartments in the city core generated more revenue on Airbnb than they likely would have on the long term market.  Examining only apartments with at least three bedrooms and two baths that figure increases to 60\%.

@@

In the logistic regression models than we estimate (Table 7), we see that most of the finding from the univariate and bivariate analyses done early are validated.  Large apartments in the central city are the best bet for generating greater revenues through short term leasing.  We also test for the impact of host behavior on the probability of short term preference.  We find that more flexibility around the number of guests and the minimum stay increase the probability of short-term preference, while the reverse is true of cancellation policies (more strict equals higher short term preference probability).  As a host, it appears to be best to do what you can to attract bookings and then make it difficult for those bookings to be cancelled.

@@

>>>>

---
title: "Should I Airbnb My Property?: \n Full Report"
author: "Andy Krause and Gideon Aschwanden"
date: "January 14, 2017"
output:
  pdf_document: default
  html_document: default
geometry: margin=3.4cm
bibliography: abb.bib
---

@@

To address these questions and provide some guidance to potential and current investors and owners, we use long-term rental data and nightly Airbnb listing information from the Melbourne (Australia) metropolitan area.  Melbourne has a thriving Airbnb market with over 20,000 total properties listed on the portal. We adopt a scenario-based approach that compares potential annual revenues for properties under both a short-term rental (STR) and a long-term rental (LTR) situation. Additionally, we break the results down by property types (apartment vs house), product type (e.g. 1 bed, 1 bath) and by location (submarket).  Our results shows that the feasibility of a greater return through a short term rental is dependent on property type, product and location.  In short, the answer to whether or not a owner or investor should opt for a short-term leasing strategy: it depends.

@@

Existing research into short-term rental markets is sparse. However, with the arrival of Airbnb, and the controversies surrounding it, the topic is creating interest both in real estate and economics as well as in a wide variety of fields as evidenced by a growth in recent publications and on-going working papers.

@@

We will not, a priori, specify a location or property type as we hypothesize that the preference of a short-term rental over a long-term one will vary by the location and property characteristics.  Additionally, Australia implements a particular tax policy, often termed `negative gearing', that does allow for favourable tax treatment to real estate investors who experience greater costs than revenues on investment properties^[See @fane2004negative for a complete explanation of negative gearing.]  However, as the negative gearing policy requires knowledge of the investor's personal income situation, we will not be considering any effects of this policy in our analysis here.

@@

We will begin our analysis by making a comparison of the short and long-term markets at a metro-level.  What we mean by this is that we will treat the entire region as the functional market and will impute rates and rents with a model using all data.  It is likely that different submarkets operate differently, a hypothesis investigated after the full metropolitan region.

@@

### Market Summary

@@

First, we bifurcate our data into houses and apartments.  We then rerun the entire analysis described above for the house and the apartment markets separately. Table 5 and Figures 18 (houses) and 19 (apartments) illustrate the metrics for these two sub-markets.  The raw probability data shows that over 33% of apartments are more profitable in short term leasing as opposed to just 22% of houses.  Figures 18 and 19 show the quantile preference plots both in the raw and with the SVM categorization.  As is evident from these plots, the likely of short-term rental being more profitable is much greater in the apartment market than in the house market.  For apartments, a 60th rate and a 60th percentile occupancy rate has a better than 50% chance of being more profitable in a short term leasing strategy than a long term one.  In fact, so long as an apartment is in the top 1/4 of occupancy rates, it is more profitable in a short term strategy at almost any price point.  Houses, on the other hand, need either very high rates or very high occupancies to justify a short-term rental.

@@

We break the data down further by geographic submarket and property type and compare the results (quantile) in Table 6.  In the Actual (observed) revenues, the probabilty of short-term preference is highest in the core area of the city and lowest in the city, beach and suburban areas.  Interestingly, the rural area has a higher probably than all other but the core.  This is likely explained by the fact that there are few rural properties and many of these are clustered near rural tourists activities like the Yarra Valley wine region which see a high demand for lodging (Figure 8).

In the core and the city, apartments offer a higher probability than homes, the reverse is true in the suburban and rural areas. In the lower density areas it appears that tourists may prefer larger and more spacious dwellings compared to long-term renters who may experience long commutes from these fringe locations.

Looking at the 'fitted' and 'svm area' figures we see a trend that the svm values are higher than the actual observed probabilities for the core apartment and lower for nearly all of the rest of the categories.  This finding suggests that for all but the core apartments, those properties that are more profitable as short-term cluster tightly in the area (the combination of nightly rate and occupancy rate) that are preferable for a short-term strategy. In other words, there is well defined combination or situation in which these properties can be more profitable short-term than in a long-term lease.  The opposite is true of the city core apartments. Here, the area of the plots occupied by short term preference is larger than the total proporation, suggesting that a wide variety of combinations of nightly rates and occupancy rates will result in a short term preference.

@@

Finally, in Model 3 we add in host behavior or policies. Allowing a greater number of guests per bedroom results in a higher probability of short-term preference.  This variable may be operating a proxy for additional sleeping spaces which allow for a higher nightly rate, ceteris paribus, to be charged.  The greater the minimum stay, the lower the probability of short-term preference.  Longer minimum stay may equate to lower occupancy rates as one or two night travelers are exempted from the property.  Owners with longer minimum stays may be looking to minimize costs such as cleaning and administration.  Flexible cancellation policies result in a lower probability of short-term preference, while a Strict policy has a weak, but positive relationship with short-term preference.  Strict policies do help maintain revenues, but may also scare off some potential guests.  Strict cancellation may be a luxury for the best and highly desired properties on the market and, within our model, may be acting as a proxy for some unobserved measure of quality or demand.

@@

<!-- Although it gathers considerable interest in both economic and public policy circles, analyses of Airbnb market and their comparison to more traditional long-term leasing strategies remains relatively unaddressed.  This work attempts to fill this void.  -->

The Airbnb market is Melbourne is rapidly expanding, as total listings on the portal have jumped from around 7,000 in 2014 to more than 20,000 in August 2016.  However, of this total, only 6,600 or so of the properties are entire homes or apartments that were listed as available for at least one day during the September 1, 2015 to August 31, 2016 time frame.  Of these, only 4,400 are common bed/bath configurations with non-outlying rates and were booked for at least one night over the period of study. Overall, the Airbnb market is very small relative to the long-term market. We caution future researchers and policy analysts to look deeply at initial estimates of Airbnb counts as many of the listings on the market are either room shares and/or not active.

@@

In looking only at the short-term observation with estimated long-term rents we found that approximately one in three Airbnb properties brought in more revenue with its short term strategy than it would have with a long-term approach. As expected, this probability is not equal across space, property type or property size (bed and bath count).  Apartments are much more likely than houses to be more profitable as a short term rental.  Core city properties are more likely than other properties to prefer a short-term strategy, as are larger properties versus smaller ones.  For instance, 45\% of apartments in the city core generated more revenue on Airbnb than they likely would have on the long term market.  Examining only apartments with at least three bedrooms and two baths that figure increases to 60\%.

@@

The logistic regression models than we estimate show that most of the finding from the univariate and bivariate analyses done earlier are validated (Table 7).  Large apartments in the central city are the best bet for generating greater revenues through short term leasing.  We also test for the impact of host behavior on the probability of short term preference.  We find that more flexibility around the number of guests and the minimum stay increase the probability of short-term preference, while the reverse is true of cancellation policies (more strict equals higher short term preference probability).  As a host, it appears to be best to do what you can to attract bookings and then make it difficult for those bookings to be cancelled.

@@

Beyond airBNB.com other short-term platforms exist (e.g. www.homeaway.com). This project doesn't take into account that individual properties may be advertised and rented out through different platforms. These are additional revenue streams that are not take into account and might increase the occupancy and revenue for short-term rentals.
